---
title: "Introduction to Data Visualisation - RStudio"
author: "Dr Peter McKenna"
date: "16/11/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=F, warning=FALSE}

 knitr::opts_chunk$set(echo = T, 
                       eval = F,    
                       message = FALSE, 
                       warning = FALSE, 
                       tidy.opts = list(width.cutoff = 60),
                       tidy = TRUE)

# default for instructor is echo = T; eval = T

# you only want to see the results for certain chunks (e.g., plots)

# add theme here to stop knitting issues

# create your own theme

library(tidyverse)

my_theme <-
  theme_classic(base_size = 15) + # change the base size of text
  theme(
    axis.title = element_text(size = rel(1.05)), # adjust the relative size of axis text
    legend.title = element_text(size = rel(1.05)), # adjust the relative size of the legend text
    plot.title = element_text(hjust = 0.5))  # centre the title (Horizontal alignment: 0 = left, 0.5 = center, 1 = right)
 
```


# **Essentials**

Before beginning today you must install two programs:

* `R`
* `RStudio`

[CLICK HERE for installation instructions for both programs by operating system.](https://www.datacamp.com/community/tutorials/installing-R-windows-mac-ubuntu)
Please follow the installation instructions carefully.

In this tutorial you will read text and code chunks.
Code chunks look like this, and contain the code for each of the operations we are going to perform today.

```{r chunk example}

# this is what a code chunk looks like

```


# DDI Data Visualisation in RStudio

As an introduction to applied data visualisation we are going to generate some plots from an existing dataset in RStudio.
I will show you how to do the following things in this exercise:

* Install & load the RStudio packages
* Create a bar plot from the `Diamonds` dataset
* Add layers to the plot to make it more informative
* Adjust the aesthetic features of the plot

**NOTE** Try not copy & paste code from this doc; typing out the code is essential to developing your programming skills!


## A little introduction

The following block of code will install and load RStudio's *Tidyverse* package.
The Tidyverse is a suit of functions and datasets that allow you to create simple and interpretable code for tidying, wrangling, visualising and analysing data.
Here, we are going to play with some of the visualising commmands of the `ggplot2` package that is loaded with the Tidyverse.


## Getting Started

To get started create a new project in RStudio and save to a meaningful location.
Once installed, open a new script (`Ctrl` + `Shift` + `N`) and enter the following code:


```{r install package, eval=F}

# Install packages

install.packages('tidyverse') # install the tidyverse package
install.packages('RColorBrewer') # install colour brewer
install.packages('plotly') # install plotly for interactive visualisations

```

Do not be concerned by RStudio's request to store packages in a local library.
This is so you can use these packages in future without the need to install.
You may also see several warning messages about the package version or conflicts. 
You can ignore these so long as the package is successfully installed. 

After a package is installed is must be loaded into RStudio for use.

```{r load package, eval=T}

# Load packages

library(tidyverse) # load the tidyverse
library(RColorBrewer) # load colour brewer
library(plotly)

```

I want to say a little about commenting here.
Using the `#` key allows you to add inactive text to your script.
So, you can add comments using `#` to keep a log of your activity


## Running the code

To run the code hit `Ctrl` + `ENTER`.
You can peform this action on multiple lines by highlighting them.
Text will appear in the console section. This is fine, RStudio is just installing the package dependencies. 
Save the script (`Ctrl` + `S`) as **ddi-vis**. 


## Download the data

The `diamonds` data is part of the Tidyverse library.
So, we can load it into our workspace using a few simple commands

```{r load data, eval=T}

data <-    # this is the assign symbol
  diamonds # we have assigned the existing dataset 'diamonds' to our dataframe 'data'

```


## The Diamonds Dataset

Here is the official [Diamonds dataset webpage](https://ggplot2.tidyverse.org/reference/diamonds.html).
Diamonds is a data frame with 53940 rows and 10 variables.
Here's information about the variables in the data:

* **price** = price in US dollars ($326--$18,823)
* **carat** = weight of the diamond (0.2--5.01)
* **cut** = quality of the cut (Fair, Good, Very Good, Premium, Ideal)
* **color** = diamond colour, from D (best) to J (worst)
* **clarity** = a measurement of how clear the diamond is (I1 (worst), SI2, SI1, VS2, VS1, VVS2, VVS1, IF (best))
* **x** = length in mm (0--10.74)
* **y** = width in mm (0--58.9)
* **z** = depth in mm (0--31.8)
* **depth** = total depth percentage = z / mean(x, y) = 2 * z / (x + y) (43--79)
* **table** = width of top of diamond relative to widest point (43--95)


## Examining the data

We can take a quick look at the data by clicking the `data` object that has been created in the Global Environment or by using the following code

``` {r Examining the data}

View(data) # opens a new tab with a large subset of the dataset

glimpse(data) # gives a summary of the data variables

head(data) # top 6 rows

tail(data) # final 6 rows

dim(data) # dimensions of the data object

```


As the dataset is quite large with several variables it's difficult to glean any trends just by observing the raw data.
This is where the power of data visualisation really comes to the fore.


# Visualising the Diamonds Dataset 

Let's start by plotting the frequency of the different cut qualities.


### Plot of Cut Quality Frequency

Add a comment using `#` to state what you are about to do.
Then enter the following code.

```{r plot 1, eval = T}

ggplot(data) +          # call ggplot2, use data dataframe "data" 
  geom_bar(mapping = aes(x = cut)) # set `cut` as our x-axis variable

```

This plot is useful in that we can see which of the cuts is most frequent in the dataset.
Most of the diamonds are an "ideal" cut.
However, it is not very informative (as it only covers one variable) nor aesthetically pleasing.
In a good plot the key information should pop out at you.


### Adding Colour

One way to make you plot pop is with colour.

```{r plot 2}

ggplot(data) +
  geom_bar(mapping = aes(x = cut, fill = cut)) # set the bar fill to `cut` variable

```

Here we get a much clearer distinction between the different cuts.


### Adding another variable

We can augment this plot with information about the diamonds clarity.

```{r plot 3}

ggplot(data) +
  geom_bar(mapping = aes(x = cut, fill = clarity)) # set the bar fill to `clarity`

```

Have a look at the variables summary above.
In the variable `clarity` I1 = low clarity; IF = high clarity.
So, you can begin to appreciate how good data visualization tells a detailed story about the data.


### Adjusting the Aesthetics: Transparency

As `clarity` is present as a variable, you may wish to adjust the bars transparency for thematic effect.

```{r plot 4}

ggplot(data, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5,   # set the alpha transparency level to 1/5
           position = "identity") # use the raw data values

```


#### Going too far

It's easy to overcook things when playing with the aesthetics.
This is what the plot looks like if you remove the fill colouring.

```{r plot 5}

ggplot(data, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(fill = NA, position = "identity")

```


### Changing Barplot Type

We can stack the variables `cut` and `clarity` to present proportions in a single
bar using the argument `position = "fill"`. 

```{r plot 6}

ggplot(data) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill") 

# `position = "fill"` tells ggplot to stack by `clarity` categories

```

So, the y-axis now represents the proportion of the dataset (e.g., from 0-1). 
Whether or not this is easier to interpret is subjective.

You can also try stacking the bars side by side using `position = "dodge"`

```{r plot 7}

ggplot(data) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge") 

# `position = "dodge"` tells ggplot to side-stack by `clarity` units

```


### Tidying Up Other Elements

#### Plot Theme

The plot theme is another important aesthetic feature.
There are several existing themes in the `ggplot2` library.

```{r plot theme}

# classic

ggplot(data) +
  geom_bar(mapping = aes(x = cut, fill = clarity)) +
  theme_classic()

# minimal

ggplot(data) +
  geom_bar(mapping = aes(x = cut, fill = clarity)) +
  theme_minimal()

# dark

ggplot(data) +
  geom_bar(mapping = aes(x = cut, fill = clarity)) +
  theme_dark()

```


#### Create Theme to Standardise Plot Aesthetic

ggplot axis label text size quite small by default
We can edit this using the `theme` command

```{r plot axes}

# create your own theme

my_theme <-
  theme_classic(base_size = 15) + # change the base size of text
  theme(
    axis.title = element_text(size = rel(1.05)), # adjust the relative size of axis text
    legend.title = element_text(size = rel(1.05)), # adjust the relative size of the legend text
    plot.title = element_text(hjust = 0.5))  # centre the title (Horizontal alignment: 0 = left, 0.5 = center, 1 = right)

# apply theme to a plot

ggplot(data) +
  geom_bar(mapping = aes(x = cut, fill = clarity)) +
  ggtitle("Cut and Clarity of Diamonds") +
  my_theme 

```


Axis labels can be modified using the `labs` command.

```{r lables}

# labels can be amended with the `labs` argument

ggplot(data) +
  geom_bar(mapping = aes(x = cut, fill = clarity)) +
  ggtitle("Cut and Clarity of Diamonds") +
  labs(x = "\nCut", y = "Frequency\n") +
  my_theme
  
```  

Notice the use of the newline `\n` command to space out the axis labels better.


## Examples of other plot types

I have given you the tools for building tidy bar-plots.
However, there are many other types of plot.
Let's stick to the `diamonds` dataset and try some others for fun.


### Jitter Plot of Diamonds Cut and Clarity

```{r jitter, eval=T}

ggplot(data, aes(cut, y = clarity)) + 
  geom_jitter(alpha = 0.05) +
  labs(title = "Diamond Frequency by Cut and Clarity", 
       x = "\nCut", 
       y = "Clarity\n") +
  my_theme

```


`geom_jitter` generates a dotplot that “jitters” the points by adding statistical noise. 
This makes it easier to compare counts between all combinations of two variables (in our case `cut` and `clarity`).

### Diamond data Boxplots

Boxplots can also be used to examine the relationship between the diamonds price cut and quality. 
As this plot will include 3 variables it can be said to be multivariate.

I've assigned the plot to an object first to apply a new colour scheme with the RColourBrewer package. 


```{r boxplot1, eval = T}

bp <-
  data %>%
  dplyr::filter(cut %in% c("Premium", "Ideal")) %>% # let's take a look at Premium and Ideal cuts spcifically
  ggplot(., aes(factor(cut), price, fill=clarity)) + 
  geom_boxplot(outlier.shape = NA) + # do not plot outliers
  ggtitle("Diamond Price According to Cut and Clarity\n") + 
  labs(x = "Cut\n", y = "\nDiamond Price US$") +
  ylim(0,7500) +
  coord_flip() # flip the plot 90 degrees clockwise
  
  

bp + 
  my_theme +
  scale_fill_brewer(palette="Set2") +
  guides(fill = guide_legend(title = "Clarity", # rename the legend
                             reverse = TRUE)) # reverse the order of the legend items

```
The bold lines on the box represent the median, and the 'whiskers' are the minimum and maximum.  


### Plotting Relationships using Facet

```{r multivariate diamonds plot}

bf <-
  data %>%
  dplyr::filter(cut %in% c("Premium", "Ideal"),
                color %in% c("D", "E", "J")) %>% # let's take a look at Premium and Ideal cuts spcifically
  ggplot(., aes(factor(cut), price, fill=clarity)) + 
  geom_boxplot() + 
  ggtitle("Diamond Price According to Colour, Clarity and Cut\n") + 
  labs(x = "Cut\n", y = "\nDiamond Price U$") + 
  ylim(0,7500) +
  coord_flip() + # flip the plot 90 degrees clockwise
  facet_grid(cols = vars(color))
  

bf +
  my_theme +
  scale_fill_brewer(palette="Set2") +
  guides(fill = guide_legend(title = "Clarity", 
                             reverse = TRUE)) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1)) # adjust x-axis text angle



```


### Creating an Interactive Plot using `plotly`

The `plotly` package allows us to add levels of interactivity to our visualisations
such as subsetting displays by selecting variable levels. 

```{r static to interactive plot with plotly}

bp <-
  ggplot(data) +
  geom_bar(mapping = aes(x = cut, fill = clarity)) +
  ggtitle("Cut and Clarity of Diamonds") +
  labs(x = "\nCut", y = "Frequency\n") +
  my_theme

ggplotly(bp) # create interactve version of plot


```


# Summary

* RStudio is a great tool for learning about data visualisation.
* There are datasets in the `tidyverse` package ideal for deploying visualisation techniques, creating exercises, and assessments. 
* Today, I've scratched the surface of what is possible, and hopefully whet your data vis appetites!
* There are plenty of free online resources dedicated to plotting in RStudio.

# Additional resources

[R Graphics Cookbook, 2nd edition](https://r-graphics.org/index.html) -- very useful starting guide for ggplot2

[R for Data Science](http://r4ds.had.co.nz/) -- useful for learning *Tidyverse* syntax

[Bookdown Chapter on Data Visualisation](https://bookdown.org/mikemahoney218/IDEAR/visualizing-your-data.html)
 

