---
title: "ddi-intro-r"
author: "petermckenna"
date: "29/11/2021"
output: html_document
---

---
title: "Introduction to Data Visualisation - RStudio"
author: "Dr Peter McKenna | p.mckenna@hw.ac.uk"
date: "16/11/2021"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
    number_sections: yes
    fig_width: 7
    fig_height: 6
    fig_caption: yes
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
---

```{r setup, include=F, warning=FALSE}

# knitr setup

knitr::opts_chunk$set(echo = T, 
                       eval = F,    
                       message = FALSE, 
                       warning = FALSE, 
                       tidy.opts = list(width.cutoff = 60),
                       tidy = TRUE)

# default for instructor is echo = T; eval = T
# you only want to see the results for certain chunks (e.g., plots)
# add theme here to stop knitting issues

# create your own theme

library(tidyverse)

my_theme <-
  theme_classic(base_size = 15) + # change the base size of text
  theme(
    axis.title = element_text(size = rel(1.05)), # adjust the relative size of axis text
    legend.title = element_text(size = rel(1.05)), # adjust the relative size of the legend text
    plot.title = element_text(hjust = 0.5))  # centre the title (Horizontal alignment: 0 = left, 0.5 = center, 1 = right)
 
```


# **Essentials**

Before beginning today you must install two programs:

* `R`
* `RStudio`

[CLICK HERE for installation instructions for both programs by operating system.](https://www.datacamp.com/community/tutorials/installing-R-windows-mac-ubuntu)
Please follow the installation instructions carefully.

In this tutorial you will read text and code chunks.
Code chunks look like this, and contain the code for each of the operations we are going to perform today.


```{r chunk example}
# this is what a code chunk looks like
```

# Introduction to R and RStudio

* `R` is a programming language
* Powerful tool for performing mathematical and statistical operations on data
* Become popular in recent years for data science due to versatility
* Today I will scratch the surface of R's functionality


# Today's intro content

* Get you a little familiar with operations of `base R`
* Walk through some programming examples
* Introduce you to the `tidyverse`: a popular data science ecosystem 
* Perform data manipulation using the `tidyverse` functions


# Introduction to R


## Create subdirectories

* Before we get started let's create some folders in our directory

```{r folder setup}

dir.create("data_raw")
dir.create("data_tidy")
dir.create("fig_output")

```

* To get started create a new project in RStudio and save to a meaningful location
* These folders will come in handy further into our exercises.


## Install and load packages

* the following packages need to be installed an loaded
* R will ask if it can create a directory to store the packages - click `OK` to continue

```{r install package, eval=F}

# Install packages

install.packages('tidyverse') # install the tidyverse package(s)

```

* Do not be concerned by RStudio's request to store packages in a local library.
* This is so you can use these packages in future without the need to install.
* You may also see several warning messages about the package version or conflicts. 
* You can ignore these so long as the package is successfully installed. 

After a package is installed is must be loaded into RStudio for use.

```{r load package, eval=T}

# Load packages

library(tidyverse) # load the tidyverse

```

* I want to say a little about commenting here.
* Using the `#` key allows you to add inactive text to your script.
* So, you can add comments using `#` to keep a log of your activity


### Running the Code

* To run the code hit `Ctrl` + `ENTER`.
* You can perform this action on multiple lines by highlighting them.
* Text will appear in the console section. This is fine, RStudio is just installing the package dependencies. 
* Save the script (`Ctrl` + `S`) as **ddi-r-intro**. 


## Some Basics

```{r basics}

# some simple operations

3+5

12/7

x <- 5 # we use the assign operator "<-" to assign values to an object

7 -> x

x

# some simple mathematics

x <- 23

y <- 67

area_rect <- x*y

# BMI

height <- 1.8

weight <- 80

bmi <- 
  (weight/(height*height))


```

## Using functions

```{r functions}

# square root

sqrt(4)

a <- 4

sqrt(a)

# rounding numeric values

round(3.14159)

args(round)

help(round)

?round

round(3.14159, digits = 2)

round(3.14159, 2)

round(digits = 2, x = 3.14159)

round(55.15, -2)

round(55.15, -3)

```


## Vectors and data types

```{r vectors and data types}

# create vector for test score

test_scores <- 
  c(3,5,2,8,7)

test_scores

# create vector for students

students <-
  c("Ronald", "Sophie", "Max", "Teresa", "Mandy")


# check length of test_score

length(test_scores)

length(students)

# check data type

class(test_scores)

class(students)

str(test_scores)

str(students)

# appending

students <-
  c(students, "Dennis") # add Dennis to the students vector

students



```

### Mixed data types

* Let's create and examine an object with both numeric and character units

```{r mixed data types}

test_answers <-
  c("a", 5, 8, TRUE, 3i)

typeof(test_answers)

class(test_answers)

str(test_answers)

test_answers2 <-
  c(FALSE, 4, 7, TRUE)

str(test_answers2)

```

### Subsetting vectors

```{r subsetting vectors}

# subsetting from our vectors

test_answers

test_answers[2] # return 2nd element

test_answers[c(3,2)] # return elements in the order you specify

students[c(5,4,2)]

# what about elements that don't exist?

students[c(9,23,4,1)]

```

### Conditional subsetting

* Let's amend the test_score vector slightly before we continue

```{r amend test_score for exercises}

# let's use the append the test_scores vector

test_scores <-
  c(test_scores, 10, 2, 1, 3, 6)

```

* Type of conditional subsetting

```{r conditional subsetting exercises}

ls() # clear the consolse

test_scores

test_scores > 3 # boolean assessment

test_scores[test_scores > 3] # subsets the vector

# using boolean operators

test_scores[test_scores <= 2 | test_scores == 6] # `|` is the OR operator

test_scores[test_scores <= 3 & test_scores > 7] # `&` is the AND operator

# boolean operators on character vector

students 

students[students == "Max" | students == "Mandy"]

# using the %in% operator

students %in% c("Teresa", "Ronald")

students[students %in% c("Teresa", "Ronald")]

```

# End of Intro to R

----

# Introduction to `tidyverse`

* In this introductory class you are going to learn how to
  + read a .csv data file into R Studio
  + wrangle and explore the data using the *Tidyverse* functions
  + explore the data's distribution
  + generate plots
  
# The `tidyverse`

* Today we are using `tidyverse` methods to explore and wrangle the data
* The Tidyverse is a set of R packages that allows for more user-friendly programming, relative to what's called *base R*
* For more about the Tidyverse see Hadley Wickham's free online text [R for Data Science](http://r4ds.had.co.nz/)
* We are covering content from R for Data Science chapters 1,2,3 & 11.


## Main verbs of R's Tidyverse

* The `tidyverse` packages allow us to use the following verbs:

    + `filter`        : extract rows
    + `select`        : extract columns 
    + `pivot_wider`   : spread rows to columns
    + `pivot_longer`  : gather columns into rows
    + `mutate`        : compute and append new/existing columns
    + `summarise`     : summarise data based on stated criteria

* These verbs make the syntax of the previous operations (e.g., subsetting) easier to interpret


## The Pipe operator

* You are going to see a lot of this symbol

```{r piping, eval=FALSE}

%>%

```

* This is the *pipe* operator
* Do not fear the pipe operator
* It means "then do this"


## Download the data

```{r download data, echo = TRUE, eval = FALSE, warning = FALSE}

# download data file from my personal github

download.file("https://github.com/petermckenna/ddi/blob/main/exp_data.csv", "data_raw/exp_data.csv", mode = "wb")


```



