---
title: "ddi-intro-r"
author: "petermckenna"
date: "29/11/2021"
output: html_document
---

---
title: "Introduction to Data Visualisation - RStudio"
author: "Dr Peter McKenna | p.mckenna@hw.ac.uk"
date: "16/11/2021"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
    number_sections: yes
    fig_width: 7
    fig_height: 6
    fig_caption: yes
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
---

```{r setup, include=F, warning=FALSE}

# knitr setup

knitr::opts_chunk$set(echo = T, 
                       eval = F,    
                       message = FALSE, 
                       warning = FALSE, 
                       tidy.opts = list(width.cutoff = 60),
                       tidy = TRUE)

# default for instructor is echo = T; eval = T
# you only want to see the results for certain chunks (e.g., plots)
# add theme here to stop knitting issues

# create your own theme

library(tidyverse)

my_theme <-
  theme_classic(base_size = 15) + # change the base size of text
  theme(
    axis.title = element_text(size = rel(1.05)), # adjust the relative size of axis text
    legend.title = element_text(size = rel(1.05)), # adjust the relative size of the legend text
    plot.title = element_text(hjust = 0.5))  # centre the title (Horizontal alignment: 0 = left, 0.5 = center, 1 = right)
 
```


# **Essentials**

Before beginning today you must install two programs:

* `R`
* `RStudio`

[CLICK HERE for installation instructions for both programs by operating system.](https://www.datacamp.com/community/tutorials/installing-R-windows-mac-ubuntu)
Please follow the installation instructions carefully.

All the code from today's session is available on my personal Github: [CLICK HERE TO ACCESS CODE](https://github.com/petermckenna/ddi/blob/main/ddi-rstudio-intro.Rmd).

In this tutorial you will read text and code chunks.
Code chunks look like this, and contain the code for each of the operations we are going to perform today.


```{r chunk example}
# this is what a code chunk looks like
```


# Introduction to R and RStudio

* `R` is a programming language
* Powerful tool for performing mathematical and statistical operations on data
* Become popular in recent years for data science due to versatility
* Today I will scratch the surface of R's functionality


# Today's intro content

* Get you a little familiar with operations of `base R`
* Walk through some programming examples
* Introduce you to the `tidyverse`: a popular data science ecosystem 
* Perform data manipulation using the `tidyverse` functions


# Introduction to R


## Create subdirectories

* Before we get started let's create some folders in our directory

```{r folder setup}

dir.create("data_raw")
dir.create("data_tidy")
dir.create("fig_output")

```

* To get started create a new project in RStudio and save to a meaningful location
* These folders will come in handy further into our exercises.


## Install and load packages

* the following packages need to be installed an loaded
* R will ask if it can create a directory to store the packages - click `OK` to continue

```{r install package, eval=F}

# Install packages

install.packages('tidyverse') # install the tidyverse package(s)

```

* Do not be concerned by RStudio's request to store packages in a local library.
* This is so you can use these packages in future without the need to install.
* You may also see several warning messages about the package version or conflicts. 
* You can ignore these so long as the package is successfully installed. 

After a package is installed is must be loaded into RStudio for use.

```{r load package, eval=T}

# Load packages

library(tidyverse) # load the tidyverse

```

* I want to say a little about commenting here.
* Using the `#` key allows you to add inactive text to your script.
* So, you can add comments using `#` to keep a log of your activity


### Running the Code

* To run the code hit `Ctrl` + `ENTER`.
* You can perform this action on multiple lines by highlighting them.
* Text will appear in the console section. This is fine, RStudio is just installing the package dependencies. 
* Save the script (`Ctrl` + `S`) as **ddi-r-intro**. 


## Some Basics

```{r basics}

# some simple operations

3+5

12/7

x <- 5 # we use the assign operator "<-" to assign values to an object

7 -> x

x

# some simple mathematics

x <- 23

y <- 67

area_rect <- x*y

# BMI

height <- 1.8

weight <- 80

bmi <- 
  (weight/(height*height))


```

## Using functions

```{r functions}

# square root

sqrt(4)

a <- 4

sqrt(a)

# rounding numeric values

round(3.14159)

args(round)

help(round)

?round

round(3.14159, digits = 2)

round(3.14159, 2)

round(digits = 2, x = 3.14159)

round(55.15, -2)

round(55.15, -3)

```


## Vectors and data types

```{r vectors and data types}

# create vector for test score

test_scores <- 
  c(3,5,2,8,7)

test_scores

# create vector for students

students <-
  c("Ronald", "Sophie", "Max", "Teresa", "Mandy")


# check length of test_score

length(test_scores)

length(students)

# check data type

class(test_scores)

class(students)

str(test_scores)

str(students)

# appending

students <-
  c(students, "Dennis") # add Dennis to the students vector

students



```

### Mixed data types

* Let's create and examine an object with both numeric and character units

```{r mixed data types}

test_answers <-
  c("a", 5, 8, TRUE, 3i)

typeof(test_answers)

class(test_answers)

str(test_answers)

test_answers2 <-
  c(FALSE, 4, 7, TRUE)

str(test_answers2)

```

### Subsetting vectors

```{r subsetting vectors}

# subsetting from our vectors

test_answers

test_answers[2] # return 2nd element

test_answers[c(3,2)] # return elements in the order you specify

students[c(5,4,2)]

# what about elements that don't exist?

students[c(9,23,4,1)]

```

### Conditional subsetting

* Let's amend the test_score vector slightly before we continue

```{r amend test_score for exercises}

# let's use the append the test_scores vector

test_scores <-
  c(test_scores, 10, 2, 1, 3, 6)

```

* Type of conditional subsetting

```{r conditional subsetting exercises}

ls() # clear the consolse

test_scores

test_scores > 3 # boolean assessment

test_scores[test_scores > 3] # subsets the vector

# using boolean operators

test_scores[test_scores <= 2 | test_scores == 6] # `|` is the OR operator

test_scores[test_scores <= 3 & test_scores > 7] # `&` is the AND operator

# boolean operators on character vector

students 

students[students == "Max" | students == "Mandy"]

# using the %in% operator

students %in% c("Teresa", "Ronald")

students[students %in% c("Teresa", "Ronald")]

```

# End of Intro to R

----

# Introduction to `tidyverse`

* In this introductory class you are going to learn how to
  + read a .csv data file into R Studio
  + wrangle and explore the data using the *Tidyverse* functions
  + explore the data's distribution
  + generate plots
  
# The `tidyverse`

* Today we are using `tidyverse` methods to explore and wrangle the data
* The Tidyverse is a set of R packages that allows for more user-friendly programming, relative to what's called *base R*
* For more about the Tidyverse see Hadley Wickham's free online text [R for Data Science](http://r4ds.had.co.nz/)
* We are covering content from R for Data Science chapters 1,2,3 & 11.


## Main verbs of R's Tidyverse

* The `tidyverse` packages allow us to use the following verbs:

    + `filter`        : extract rows
    + `select`        : extract columns 
    + `pivot_wider`   : spread rows to columns
    + `pivot_longer`  : gather columns into rows
    + `mutate`        : compute and append new/existing columns
    + `summarise`     : summarise data based on stated criteria

* These verbs make the syntax of the previous operations (e.g., subsetting) easier to interpret


## The Pipe operator

* You are going to see a lot of this symbol

```{r piping, eval=FALSE}

%>%

```

* This is the *pipe* operator
* Do not fear the pipe operator
* It means "then do this"


# Working with data objects

## Download the data

```{r download data, echo = TRUE, eval = FALSE, warning = FALSE}

# download data file from my personal github

download.file("https://github.com/petermckenna/ddi/blob/main/exp_data.csv", "data_raw/exp_data.csv", mode = "wb")


```

## Read in the data

* Now that we have downloaded the data we nened to read it in
* Today we are going to read in a .csv file
* RStudio can read in all data types with the package `readr` (e.g., .xls)

```{r reading in the data}

data <-                      # assign ( <- ) the label "data" to next set of commands
  read_csv("exp_data.csv") # read in csv file called "exp_data.csv"

```


## Experiment data info

* `id`         = participant number
* `like`       = rating of assistants likeability from 1 = "Did not like at all" to 5 = "Liked it a lot" 
* `voice_type` = speaker voice type including two levels: Female; Male. 
* `conv_len`   = duration of the conversation between the user and the smart speaker measured in sec


## Inspecting the data

* Let's go through each of the commands below 
* Checking data with these commands helps you to get a better understanding of the data
* `head` & `tail` for example help you quickly determine if the dataset has been read in properly from top to bottom 


```{r inspecting the data, echo = TRUE, eval = FALSE}

head(data)    # shows top 6 rows

tail(data)    # shows bottom 6 rows

dim(data)     # number of rows and columns

names(data)   # list vector names

# you can use square brackets to give an index to certain elements of the data

names(data)[2] # give the name of column 2

View(data)    # opens subset of the data in a new tab; note the capital 'V'

glimpse(data) # short by-vector summary

```


# Data wrangling and manipulation with the `tidyverse`

## Using `filter` to subset by row

```{r filtering data}

# filtering by single element

data %>%
  filter(id == 1:5) # participants 1 to 5

data %>%
  filter(voice_type != "female") # "!=" denotes exclude

# filtering by multiple elements

data %>%
  filter(id == 1:5, 
         voice_type == "female")      

```

## Using `mutate` to append object

```{r mutate data}

# say we knew that participants 1:6 were female and the rest were male

# we can use an ifelse statement to create a new vector `gender` based on this criteria

data %>%
  mutate(gender = ifelse(id %in% 1:6, 
                         "female", 
                         "male"))

```

## Using `select` to subset by column

```{r select subset by col}

data %>%
  select(id, like) # you can use a colon to select multiple vectors 

# using do not select

data %>%
  select(-id) # everything except `id`

# you can also reorder existing cols with select

data %>%
  select(like, conv_len, id)

# or with vector numbers instead of names

data %>%
  select(4, 2, 3, 1)

```

# Summarising data

* In research, we are interested in summary statistics
* I'll give you a brief overview of how to generate using the `tidyverse`

```{r summarising data}

data %>%
  summarise(med_like = median(like)) # here we generate the median of `like` and call the summary "med_like"


data %>%
  summarise(med_rating = median(like),    
            avg_conv = mean(conv_len),       # new vector `avg_conv` is the mean of conv_len
            sd_conv = sd(conv_len))          # new vector `sd_conv` is S.D. of conv_len

# grouping data

  data %>%                               # create new object `bp` from data
  group_by(voice_type) %>%               # group by `voice_type`                    
  summarise(avg_conv = mean(conv_len),   # new vector avg_conv is the mean of conv_len 
            sd_conv = sd(conv_len))      # new vector sd_conv is the S.D. of conv_len


```
















